---
title: "Map GWAS to VCF workflow"
author: "Matt Lyon"
date: "2020-02-25"
params:
    n_sim: 5
---

### Download and format GWAS

Download BMI GWAS summary statistics and annotation data (Neale et al)

```{r}
library('data.table')
library('stringr')
library('jsonlite')
set.seed(12345)
options(scipen=999)

# UK Biobank — Neale lab. (n.d.). Retrieved February 25, 2020, from http://www.nealelab.is/uk-biobank/
download.file("https://www.dropbox.com/s/tx8iw9lk53z3vj9/21001_raw.gwas.imputed_v3.both_sexes.tsv.bgz?raw=1", destfile="/data/gwas.raw.txt.gz")
download.file("https://www.dropbox.com/s/puxks683vb0omeg/variants.tsv.bgz?raw=1", destfile="/data/lookup.txt.gz")
gwas <- fread("/data/gwas.raw.txt.gz")
lookup <- fread("/data/lookup.txt.gz")
gwas <- merge(gwas, lookup, "variant")
rm(lookup)

# save output to gzip tab file
df1 <- gwas[,c("chr", "pos", "alt", "ref", "beta", "se", "pval", "rsid", "AF", "info", "n_complete_samples")]
gz1 <- gzfile("/data/gwas.raw.merged.txt.gz", "w")
write.table(df1, gz1, sep="\t", row.names=F, quote=F)
close(gz1)

# write out params to json
j <- list(chr_col = 0, pos_col = 1, ea_col = 2, oa_col = 3, beta_col = 4, se_col = 5, pval_col = 6, snp_col = 7, eaf_col = 8, imp_info_col = 9, ncontrol_col = 10, delimiter = "\t", header = TRUE, build = "GRCh37")
write(toJSON(j, auto_unbox=T), file = "/data/schema.json")
```

### Map GWAS to VCF

Map plain text to VCF format

Convert GWAS to VCF. This will perform on-the-fly harmonisation of the effect allele to the VCF alternative allele ensuring data consistency. Source code available here: [gwas2vcf](https://github.com/MRCIEU/gwas2vcf).

```{bash}
set -euo pipefail
python /app/gwas2vcf/main.py \
--out /data/gwas.raw.vcf.gz \
--data /data/gwas.raw.merged.txt.gz \
--ref /data/human_g1k_v37.fasta \
--id "neale-bmi-ukb" \
--json /data/schema.json
```

### Combine multialleleic variants

Merge duplicate positions into a single row. This is required if using ID field to prevent duplicate identifiers.

```{bash}
set -euo pipefail
bcftools norm \
-f /data/human_g1k_v37.fasta \
-m +any \
-O z \
-o /data/gwas.norm.vcf.gz \
/data/gwas.raw.vcf.gz
```

### Update ID field

Replace ID field with up-to-date set of dbSNP identifiers using versioned release. IDs are frequently merged between builds. Also by combining multialleleics rsIDs for INDELs can reference the wrong position.

Strip out existing ID field

```{bash}
set -euo pipefail
bcftools annotate \
-x ID \
-O z \
-o /data/gwas.noid.vcf.gz \
/data/gwas.norm.vcf.gz

bcftools index \
-t \
/data/gwas.noid.vcf.gz
```

Add versioned IDs (build 153)

```{bash}
set -euo pipefail
bcftools annotate \
-c ID \
-a /data/dbsnp.v153.b37.vcf.gz \
-O z \
-o /data/gwas.all.vcf.gz \
/data/gwas.noid.vcf.gz

bcftools index \
-t \
/data/gwas.all.vcf.gz
```

### Drop multialleleics and missing rsID field

Not including these variants in the performance comparison for simplicity

```{bash}
bcftools view \
-e 'COUNT(ALT) > 1' \
-e 'ID == "."' \
-O z \
-o /data/gwas.vcf.gz \
/data/gwas.all.vcf.gz

bcftools index \
-t \
/data/gwas.vcf.gz
```

### Validate output

Check file format is valid. Genotype checking is skipped since the file has no genotypes.

```{bash}
set -euo pipefail
/usr/bin/gatk-4.1.5.0/gatk \
ValidateVariants \
-R /data/human_g1k_v37.fasta \
-V /data/gwas.vcf.gz \
--validation-type-to-exclude ALLELES
```

### Index on rsID

[rsidx](https://github.com/bioforensics/rsidx) is an excellent tool for indexing on VCF ID. This is made possible by extracting rsID and chromosome position to an SQLite database which is later queried to obtain the variant locus for extraction using tabix.

```{bash}
rsidx index \
/data/gwas.vcf.gz \
/data/gwas.vcf.gz.rsidx
```

### Create uncompressed VCF & index

Uncompress the file to see what the performance impact is later.

```{bash}
zcat /data/gwas.vcf.gz > /data/gwas.vcf
/usr/bin/gatk-4.1.5.0/gatk \
IndexFeatureFile \
-I /data/gwas.vcf
```

### Prepare tabular extract for performance comparisons

Create extract of VCF to text for performance comparisons later

```{bash}
echo -e "CHROM\tPOS\tREF\tALT\tID\tES\tSE\tLP\tAF\tSS" > /data/gwas.txt
bcftools query \
-f '%CHROM\t%POS\t%REF\t%ALT\t%ID[\t%ES\t%SE\t%LP\t%AF\t%SS]\n' \
/data/gwas.vcf.gz >> /data/gwas.txt
gzip -c /data/gwas.txt > /data/gwas.txt.gz
```

### Define queries

```{r}
library('data.table')
set.seed(12345)
options(scipen=999)

# read in tab GWAS
gwas <- fread("/data/gwas.txt")

# subsample variants
s <- gwas[sample(nrow(gwas), params$n_sim),]

# write out list of rsid for querying later
write.table(s$ID, sep="\t", row.names=F, quote=F, col.names=F, file="/data/rsid.txt")

# write out list of chrom pos for querying later
write.table(s[,c("CHROM", "POS")], sep="\t", row.names=F, quote=F, col.names=F, file="/data/chrpos.txt")

# write out list of 1Mb intervals and count number of variants for querying later
s$start <- s$POS - 500000
s$end <- s$POS + 500000
s$interval_count <- apply(s, 1, function(row){ length(which(gwas$CHROM == row[['CHROM']] & gwas$POS > as.numeric(row[['start']]) & gwas$POS < as.numeric(row[['end']]))) })
write.table(s[,c("CHROM", "start", "end", "interval_count")], sep="\t", row.names=F, quote=F, col.names=F, file="/data/intervals.txt")

# write out pval thresholds & counts for querying later
gwas$LP <- as.numeric(gwas$LP)

## define pval thresholds
thresholds <- seq(0.00000005, 1, 1/params$n_sim)
thresholds <- -log10(thresholds)

## round to avoid float imprecision issues
thresholds <- round(thresholds, 4)

## count records
counts <- sapply(thresholds, function (threshold) {sum(gwas$LP > threshold, na.rm = T)})
write.table(data.frame(thresholds=thresholds, counts=counts), sep="\t", row.names=F, quote=F, col.names=F, file="/data/pval.txt")

message(paste("Total number of variants in GWAS:", nrow(gwas)))
```
